---
title: "snRNAseq workflow: inferCNV"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
params:
  sample: false
  rds_file: false
  gene_order_file: false
  cache_dir: false
output:
  html_document:
    keep_md: true
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    number_sections: true
    theme: lumen
---

```{r initialise, include = F}
library(magrittr)

# set seed for repeatability
set.seed(42)
random_seed = 42

# chunk setup
knitr::opts_chunk$set(fig.align = "center",
                      dpi = 300)
```

# Load object

```{r load_seu}
seu <- readRDS(params$rds_file)
```

# Copy number analysis

## Run `inferCNV`

From [**github.com/broadinstitute/inferCNV/wiki**](https://github.com/broadinstitute/inferCNV/wiki)

>  ***InferCNV: Inferring copy number alterations from tumor single cell RNA-Seq data***
>
> *InferCNV is used to explore tumor single cell RNA-Seq data to identify evidence for somatic large-scale chromosomal copy number alterations, such as gains or deletions of entire chromosomes or large segments of chromosomes. This is done by exploring expression intensity of genes across positions of tumor genome in comparison to a set of reference 'normal' cells. A heatmap is generated illustrating the relative expression intensities across each chromosome, and it often becomes readily apparent as to which regions of the tumor genome are over-abundant or less-abundant as compared to that of normal cells.*

We annotate all matched normal and use this to infer copy number variants in the malignant population. Unlike CNVs called from bulk genotyping data, `inferCNV` can also reveal subclonal events in the tumour that only occur in some cell lineages, but not others. 

```{r get_ref_cells}
ref_celltypes <- c("T_cells", "Macrophage", "B_cells", "NK_cells", "Myeloid_cells", "Stromal_cells")

# different reference groupings
# -> sample_celltype - each reference celltype is a different reference group
# -> sample_pool - all reference celltypes are pooled as a reference group
# -> patient_celltype - each reference celltype is pooled from all T0/T1/T2 samples of the patient
# -> patient_pool - all reference celltypes from all T0/T1/T2 samples of the patient are pooled 

cells <-
  seu@meta.data %>%
  tibble::as_tibble(rownames = "cell") %>%
  dplyr::select(sample, cell, celltype = label.reduced) 

query_cells <-
  cells %>%
  dplyr::filter(celltype == "Epithelial_cells" & sample == params$sample) %>%
  dplyr::select(cell, infercnv_population = celltype)

reference_cells <-
  cells %>%
  dplyr::filter(celltype %in% ref_celltypes) %>%
  dplyr::mutate(sample_celltype = dplyr::case_when(sample == params$sample ~ celltype, .default = NA),
                sample_pool = dplyr::case_when(sample == params$sample ~ "Sample_reference_cells", .default = NA),
                patient_celltype = paste0("Patient_", celltype),
                patient_pool = "Patient_reference_cells") %>%
  tidyr::pivot_longer(cols = c("sample_celltype", "sample_pool", "patient_celltype", "patient_pool"),
                      values_drop_na = T) %>%
  dplyr::group_by(name, value) %>%
  dplyr::filter(dplyr::n() > 100) %>%
  dplyr::ungroup() 

# decide reference pooling behaviour
if (nrow(query_cells) < 100) {
  message("insufficient query cells")
  final_name <- NULL
} else {
  message("sufficient query cells")
  if ("sample_celltype" %in% reference_cells$name) {
    message("sufficient reference cells from individual celltypes")
    message()
    final_name <- "sample_celltype" 
  } else if ("sample_pool" %in% reference_cells$name) {
    message("insufficient cells from individual reference celltypes")
    message("sufficient cells across all reference celltypes in the sample")
    message("must pool reference celltypes from the sample together")
    final_name <- "sample_pool"
  } else if ("patient_celltype" %in% reference_cells$name) {
    message("insufficient cells from individual reference celltypes")
    message("insufficient cells across all reference celltypes in the sample")
    message("must pool each reference celltype from the patient together")
    final_name <- "patient_celltype"
  } else if ("patient_pool" %in% reference_cells$name) {
    message("insufficient cells from individual reference celltypes")
    message("insufficient cells across all reference celltypes in the sample")
    message("insufficient cells from individual reference celltypes pooled across the patient")
    message("must pool all reference celltypes from the patient together")
    final_name <- "patient_pool"
  } else {
    message("insufficient reference cells across the patient")
    final_name <- NULL
  }
}
```

```{r final_ref_cells, eval = !is.null(final_name) == T}
# get annots
reference_cells <- 
  reference_cells %>% 
  dplyr::filter(name == final_name) %>%
  dplyr::select(cell, infercnv_population = value)
infercnv_annots <-
  reference_cells %>%
  dplyr::bind_rows(query_cells)

# view refs
infercnv_annots %>% 
  dplyr::count(infercnv_population) %>%
  knitr::kable()
```

```{r run_infercnv, eval = !is.null(final_name) == T}
# create output dir
dir.create(params$cache_dir, recursive = T, showWarnings = F)

# save annots file
infercnv_annots %>%
  readr::write_tsv(paste0(params$cache_dir, "/infercnv_annots.tsv"), col_names = F)

# subset seu_i
seu_i <- seu[, infercnv_annots$cell]

# create infercnv object
infercnv_obj <- 
  infercnv::CreateInfercnvObject(
    raw_counts_matrix = Seurat::GetAssayData(seu_i, slot = "counts"),
    annotations_file = paste0(params$cache_dir, "/infercnv_annots.tsv"),
    gene_order_file = params$gene_order_file,
    ref_group_names = unique(reference_cells$infercnv_population)
  )

# perform infercnv operations to reveal cnv signal
options(scipen = 100)
infercnv_obj <- 
 infercnv::run(
   infercnv_obj,
   cutoff = 0.1,
   out_dir = params$cache_dir,
   cluster_by_groups = T,
   denoise = T,
   HMM = T, 
   HMM_type = "i3",
   HMM_report_by = "cell",
   num_threads = 10,
   no_prelim_plot = T,
   analysis_mode = "subclusters",
   resume_mode = T
 )
```

Malignant vs non-malignant cell types are then assigned.

```{r malig_classification, eval = !is.null(final_name) == T}
# add infercnv metadata to seu, save
seu_i <- infercnv::add_to_seurat(seu_i, infercnv_output_path = params$cache_dir)

# define malignancy by aneuploidy status
cnv_file <- "HMM_CNV_predictions.HMMi3.leiden.hmm_mode-subclusters.Pnorm_0.5.pred_cnv_regions.dat"
aneuploid_cells <-
  readr::read_tsv(paste0(params$cache_dir, cnv_file)) %>%
  dplyr::group_by(cell_group_name) %>%
  dplyr::summarise(aneuploid = any(c(1, 3) %in% state)) %>%
  dplyr::filter(
    aneuploid & cell_group_name %in% colnames(seu_i[, seu_i$label.reduced == "Epithelial_cells"])) %>%
  dplyr::pull(cell_group_name)

# add malignant label to aneuploid cells
seu_i@meta.data$label.infercnv <- seu_i@meta.data$label.reduced
seu_i@meta.data[aneuploid_cells, "label.infercnv"] <- "Malignant_cells"

# save
saveRDS(seu_i, "seu_infercnv.rds")

if (length(aneuploid_cells) > 0) {
  # save malignant cells only
  seu_i_malig <- seu_i[, seu_i@meta.data$label.infercnv == "Malignant_cells"]
  saveRDS(seu_i_malig, "seu_infercnv_malignant.rds")
}
```
