---
title: "snRNAseq workflow: inferCNV"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
params:
  params_file: false
  rds_file: false
  cache_dir: false
output:
  html_document:
    keep_md: true
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    number_sections: true
    theme: lumen
---

```{r initialise, include = F}
library(magrittr)

# load params file
curr_params <- 
  rjson::fromJSON(file = params$params_file) %>% 
  type.convert(as.is = T)

# set seed for repeatability
set.seed(42)
random_seed = 42

# chunk setup
knitr::opts_chunk$set(fig.align = "center",
                      dpi = 300)
```

```{r define_functions, include = F}
# flatten lists
flatten_list <- function (x, use.names = TRUE, classes = "ANY") 
{
    len <- sum(rapply(x, function(x) 1L, classes = classes))
    y <- vector("list", len)
    i <- 0L
    items <- rapply(x, function(x) {
        i <<- i + 1L
        y[[i]] <<- x
        TRUE
    }, classes = classes)
    if (use.names && !is.null(nm <- names(items))) 
        names(y) <- nm
    y
}
```

# Run info {-}

```{r curr_params, echo = F}
curr_params %>%
  flatten_list() %>%
  {tibble::tibble(
    param = names(.), 
    value = purrr::map(., paste, collapse = ", ") %>% unlist() %>% substr(1, 250))
  } %>%
  knitr::kable()
```

# Load object

```{r load_seu}
seu <- readRDS(params$rds_file)
```

# Copy number analysis

## Run `inferCNV`

From [**github.com/broadinstitute/inferCNV/wiki**](https://github.com/broadinstitute/inferCNV/wiki)

>  ***InferCNV: Inferring copy number alterations from tumor single cell RNA-Seq data***
>
> *InferCNV is used to explore tumor single cell RNA-Seq data to identify evidence for somatic large-scale chromosomal copy number alterations, such as gains or deletions of entire chromosomes or large segments of chromosomes. This is done by exploring expression intensity of genes across positions of tumor genome in comparison to a set of reference 'normal' cells. A heatmap is generated illustrating the relative expression intensities across each chromosome, and it often becomes readily apparent as to which regions of the tumor genome are over-abundant or less-abundant as compared to that of normal cells.*

We annotate all matched normal and use this to infer copy number variants in the malignant population. Unlike CNVs called from bulk genotyping data, `inferCNV` can also reveal subclonal events in the tumour that only occur in some cell lineages, but not others. 

```{r run_infercnv}
# ref_annots <- c("T_cells", "Macrophage", "B_cells", "NK_cells", "Myeloid_cells", "Stromal_cells")
# query_annots <- "Epithelial_cells"
annots_col <- "label.reduced"
ref_annots <- stringr::str_split(curr_params$infercnv$reference_celltypes, ",")[[1]]
query_annots <- stringr::str_split(curr_params$infercnv$query_celltypes, ",")[[1]]

for (sample in unique(seu@meta.data$sample)) {
  print(sample)
  
  # create output dirs
  out_dir <- paste0(params$cache_dir, "/", sample, "/")
  dir.create(out_dir, recursive = T, showWarnings = F)
  dir.create(sample)
  
  # subset seu
  seu_i <- seu[, seu@meta.data$sample == sample]
  
  # label query and ref populations
  infercnv_annots <- 
    seu_i@meta.data %>% 
    tibble::as_tibble(rownames = "cell") %>% 
    dplyr::select(sample, cell, annot = dplyr::matches(annots_col)) %>%
    dplyr::filter(annot %in% c(ref_annots, query_annots)) %>%
    dplyr::transmute(
      cell,
      infercnv_population = dplyr::case_when(
        annot %in% query_annots ~ paste0("malignant_", sample),
        annot %in% ref_annots ~ annot
      )
    ) %>%
    # remove annotations containing <100 cells
    dplyr::group_by(infercnv_population) %>%
    dplyr::filter(dplyr::n() > 100)
  
  # save annots file
  infercnv_annots %>%
    readr::write_tsv(paste0(sample, "/infercnv_annots.tsv"), col_names = F)
  
  # subset seu_i
  seu_i <- seu_i[, seu_i@meta.data[, annots_col] %in% c(infercnv_annots$infercnv_population, query_annots)]
  
  # get filtered reference annots
  filt_ref_annots <-
    infercnv_annots %>%
    dplyr::filter(!grepl("malignant", infercnv_population)) %>%
    dplyr::pull(infercnv_population) %>%
    unique()
  
  # do not proceed if no query cells left
  if (!any(grepl("malignant", infercnv_annots$infercnv_population))) {
    message("<100 query cells. Cannot proceed.")
  } else {
    
    # create infercnv object
    infercnv_obj <- 
      infercnv::CreateInfercnvObject(
        raw_counts_matrix = Seurat::GetAssayData(seu_i, slot="counts"),
        annotations_file = paste0(sample, "/infercnv_annots.tsv"),
        gene_order_file = curr_params$infercnv$gene_order_file,
        ref_group_names = filt_ref_annots
      )
    
    # perform infercnv operations to reveal cnv signal
    options(scipen = 100)
    infercnv_obj <- 
     infercnv::run(
       infercnv_obj,
       cutoff = 0.1,
       out_dir = out_dir,
       cluster_by_groups = T,
       denoise = T,
       HMM = T, 
       HMM_type = "i3",
       HMM_report_by = "cell",
       num_threads = 10,
       no_prelim_plot = T,
       analysis_mode = "subclusters",
       resume_mode = T
     )
    
  }
}
```

Malignant vs non-malignant cell types are then assigned.

```{r malig_classification}
aneuploid_cells <-
  list.files(
    params$cache_dir, 
    recursive = T, 
    pattern = "Pnorm_0.5.pred_cnv_regions.dat", 
    full.names = T
  ) %>%
  purrr::map(function(file) {
    readr::read_tsv(file) %>%
      dplyr::mutate(sample = stringr::str_split(file, "/")[[1]][15])
  }) %>%
  dplyr::bind_rows() %>%
  dplyr::group_by(sample, cell_group_name) %>%
  dplyr::summarise(aneuploid = any(c(1, 3) %in% state)) %>%
  dplyr::filter(aneuploid) %>%
  dplyr::pull(cell_group_name)

# add malignant label to aneuploid cells
seu@meta.data$label.infercnv <- seu@meta.data$label.reduced
seu@meta.data[aneuploid_cells, "label.infercnv"] <- "Malignant_cells"

# save
saveRDS(seu, "seu_infercnv.rds")
```
