---
title: "snRNAseq workflow: dimensionality reduction and clustering"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
params:
  params_file: false
  rds_file: false
  cache_dir: false
output:
  html_document:
    keep_md: true
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    number_sections: true
    theme: lumen
---

```{r initialise, include = F}
# magrittr, gridExtra, clustree, monocle3 must be loaded in the environment (due to package-specific bugs)
library(clustree)
library(gridExtra)
library(monocle3)
library(magrittr)

# load params file
curr_params <- 
  rjson::fromJSON(file = params$params_file) %>% 
  type.convert(as.is = T)

# set seed for repeatability
set.seed(42)
random_seed = 42

# set ggplot presets
ggplot2::theme_set(ggplot2::theme_bw())
ggplot2::theme_update(legend.text = ggplot2::element_text(size = 7))
ditto_colours <- list(ggplot2::scale_fill_manual(values = dittoSeq::dittoColors(), na.value = "white"),
                      ggplot2::scale_colour_manual(values = dittoSeq::dittoColors(), na.value = "white"))
umap_void_theme <-
  ggplot2::theme(axis.text = ggplot2::element_blank(),
                 axis.title = ggplot2::element_blank(),
                 axis.ticks = ggplot2::element_blank(),
                 legend.position = "none",
                 plot.margin = ggplot2::unit(c(2,2,2,2), "pt"))

# chunk setup
knitr::opts_chunk$set(fig.align = "center",
                      dpi = 300,
                      cache.path = params$cache_dir)

# reductions
redus <- c("PCA", "UMAP")

# calculate plot height for grids
get_fig_dims <- function(n_plots, n_cols = NULL, grid_width = 10, height_to_width_ratio = 1) {
  # get dims
  dims <- ggplot2::wrap_dims(n_plots, n_cols)
  # get scale
  scale <- grid_width / dims[1]
  # get height
  grid_height <- (dims[2] * scale) * height_to_width_ratio
  dims <- c(grid_width, grid_height)
  # if any dim is 0, make 1 to avoid error
  dims[dims == 0] <- 1
  # return dims
  dims
}

# dev.off() if there is a device
dev_off_if <- function() {if (length(dev.list()) != 0) { dev.off() }}

# Get cluster centroids from a Seurat or cell_data_set object
get_centroids <- function(object, reduction, lvl) {

  if (class(object)[1] == "Seurat") {
    embeddings <- object@reductions[[reduction]]@cell.embeddings
    metadata <- object@meta.data
  } else if (class(object)[1] == "cell_data_set") {
    embeddings <- SingleCellExperiment::reducedDims(object)[[reduction]]
    metadata <- SummarizedExperiment::colData(object) %>% dplyr::as_tibble()
  }

  dplyr::tibble(
    x = embeddings[,1],
    y = embeddings[,2],
    metadata
  ) %>%
    dplyr::group_by(get(lvl)) %>%
    dplyr::summarise(x = median(x), y = median(y)) %>%
    dplyr::rename(!!lvl := `get(lvl)`)

}
```

```{r load_seu}
seu <- readRDS(params$rds_file)
```

# Dimensionality reduction and clustering

## Linear dimensionality reduction

```{r lin_dim_red, echo = F}
seu <- xfun::cache_rds({
  seu <- Seurat::FindVariableFeatures(seu)
  seu <- Seurat::SCTransform(seu)
  seu <- Seurat::CellCycleScoring(
    seu,
    s.features = Seurat::cc.genes$s.genes,
    g2m.features = Seurat::cc.genes$g2m.genes
  )
  seu
}, file = "seu_sct.rds")
seu <- xfun::cache_rds({
  seu <- Seurat::RunPCA(seu, npcs = ifelse(ncol(seu) <= 50, ncol(seu) - 1, 50))
  seu
}, file = "seu_pca.rds", rerun = T)
```

## Dimensionality of the data

```{r usr_n_dims_msg, echo = F, eval = curr_params$cluster$n_dims != F, results = 'asis'}
cat("### Manually\n")
cat(paste0("Dimensionality of the data is **", curr_params$cluster$n_dims, "** (set by argument n_dims).\n"))
```

```{r usr_n_dims, include = curr_params$cluster$n_dims != F, eval = curr_params$cluster$n_dims != F}
final_n_dims <- curr_params$cluster$n_dims
```

### intrinsicDimension

```{r iD_msg, echo = F, eval = curr_params$cluster$n_dims == F, results = 'asis'}
cat("No value was provided for `n_dims`, so dimensionality will be calculated using the intrinsicDimensions package.\n")
```

The dimensionality of the dataset can be estimated using the `intrinsicDimension` package. This bypasses the manual approach of reading the elbow plot, but can sometimes be a bit too conservative. Therefore, if `n_dims` is not defined by the user, the dimensionality is set as 1.5 x `intrinsicDimension` estimate. 

```{r iD_n_dims}
iD_n_dims <- xfun::cache_rds({
  intrinsicDimension::maxLikGlobalDimEst(
    seu@reductions$pca@cell.embeddings,
    k = 10
  )$dim.est %>% round(0)
}, file = "iD_n_dims.rds")
generous_n_dims <- round(1.5 * ifelse(iD_n_dims == 0, 1, iD_n_dims), 0)
if (generous_n_dims > ncol(seu)) { generous_n_dims <- ncol(seu) / 2 }
```

```{r iD_n_dims_set, include = curr_params$cluster$n_dims == FALSE, eval = curr_params$cluster$n_dims == FALSE}
final_n_dims <- generous_n_dims
```

Dimensionality set to **`r final_n_dims`**.

```{r elbow_plot, class.source = 'fold-hide', fig.cap = "Elbow plot"}
p <-
  Seurat::ElbowPlot(seu, ndims = max(50, final_n_dims)) &
  ggplot2::geom_vline(
    ggplot2::aes(xintercept = final_n_dims,
                 colour = ifelse(curr_params$cluster$n_dims == F, "chosen n_dims", "user-provided n_dims")),
                 linewidth = 1) &
  ggplot2::geom_vline(ggplot2::aes(xintercept = iD_n_dims, colour = "intrisicDimensions"),
                      linewidth = 1, linetype = "dashed") &
  ggplot2::geom_vline(ggplot2::aes(xintercept = generous_n_dims, colour = "generous (iD x 1.5)"),
                      linewidth = 1, linetype = "dashed") &
  ggplot2::scale_colour_manual(
    name = "n_dims cut-off",
    values = c(`chosen n_dims` = dittoSeq::dittoColors()[[1]],
               `user-provided n_dims` = dittoSeq::dittoColors()[[1]],
               intrisicDimensions = dittoSeq::dittoColors()[[2]],
               `generous (iD x 2)` = dittoSeq::dittoColors()[[3]])) &
  ggplot2::theme(legend.position = c(0.7, 0.9),
                 legend.background = ggplot2::element_rect(fill = "white"))
p
```

## Non-linear dimensionality reduction and clustering

```{r clustering_and_umap, echo = F}
clustering_resolutions <- seq(0.1, 0.8, by = 0.1)
seu <- xfun::cache_rds({
  seu <- Seurat::FindNeighbors(seu, dims = 1:final_n_dims)
  seu <- Seurat::FindClusters(seu, resolution = clustering_resolutions)
  seu <- Seurat::RunUMAP(seu, dims = 1:final_n_dims)
  seu
}, file = "seu_clustered.rds")
```

### Groupings vs reductions

```{r get_groupings, include = F, echo = F}
# sample_metadata columns with >1 value that are not `dir` or `file_prefix`
groupings <- 
  seu@misc$sample_metadata %>%
  dplyr::select(where(~ dplyr::n_distinct(.x) > 1)) %>%
  dplyr::select(-tidyr::any_of(c("dir", "file_prefix"))) %>%
  colnames()
```

```{r groupings_vs_reductions, echo = F, fig.dim = get_fig_dims(n_plots = 3 * length(groupings), n_cols = 3), fig.cap = "Plot of all monocle3 dimensionality reductions, coloured by groupings", eval = (nrow(seu@misc$sample_metadata) > 1) == T}
# plot all groupings vs all reductions
groupings_vs_reductions <- list()
purrr::walk(groupings, function(grouping) {
  purrr::walk(redus, function(redu) {
    if (dplyr::n_distinct(seu@meta.data[, grouping]) > 1) {
      title <- paste0(redu, "_vs_", grouping)
      p <- 
        Seurat::DimPlot(
          seu, 
          group.by = grouping, 
          reduction = tolower(redu),
          cols = dittoSeq::dittoColors(),
          raster = F
        )
      groupings_vs_reductions[[paste0(grouping, "_legend")]] <<-
        lemon::g_legend(p)
      groupings_vs_reductions[[title]] <<- 
        p & Seurat::NoLegend()
    }
  })
})
# create grob layout
p <-
  gridExtra::marrangeGrob(
    grobs = groupings_vs_reductions,
    nrow = length(groupings),
    ncol = length(redus) + 1,
    layout_matrix = matrix(
      1:length(groupings_vs_reductions),
      length(groupings),
      length(redus) + 1,
      TRUE
    ),
    top = NULL
  )
p
```

### Resolutions vs reductions

```{r resolutions_vs_reductions, echo = F, fig.dim = c(10, 40)}
resolutions_vs_reductions <- list()
purrr::walk(clustering_resolutions, function(res) {
  purrr::walk(Seurat::Reductions(seu), function(redu) {
      title <- paste0(redu, "_by_cluster_res_", res)
      resolutions_vs_reductions [[title]] <<-
        dittoSeq::dittoDimPlot(
          seu,
          paste0(Seurat::DefaultAssay(seu), "_snn_res.", res),
          reduction.use = redu,
          size = 0.5,
          raster = F,
          legend.show = F,
          show.axes.numbers = F,
          show.grid.lines = F,
          main = gsub("_", " ", title)
        )
    })
})
# create grob layout
p <-
  gridExtra::marrangeGrob(
    grobs = resolutions_vs_reductions ,
    nrow = length(clustering_resolutions),
    ncol = length(Seurat::Reductions(seu)),
    layout_matrix = matrix(1:length(resolutions_vs_reductions ),
                           length(clustering_resolutions),
                           length(Seurat::Reductions(seu)),
                           TRUE),
    top = NULL
  )
p
```

### Clustering tree

```{r clustering_tree, echo = F, fig.cap = "Plot of clustering tree"}
snn_res_prefixes <- paste0(Seurat::DefaultAssay(seu), "_snn_res.")
p <- 
  clustree::clustree(seu@meta.data[, grep(snn_res_prefixes, colnames(seu@meta.data))],
                       prefix = snn_res_prefixes)
p
```

### Final clustering resolution

```{r final_clusters, eval = curr_params$clustering_resolution != F}
seu$cluster <- seu@meta.data[, paste0("SCT_snn_res.", curr_params$clustering_resolution)]
seu$cluster <- factor(seu$cluster, levels = paste(sort(as.integer(levels(seu$cluster)))))
Seurat::Idents(seu) <- seu$cluster
cluster_metadata <- xfun::cache_rds({
  seu@meta.data
}, file = "cluster_metadata.rds")

p <- Seurat::DimPlot(seu)
p
```

#### Final clustering resolution markers

```{r cluster_markers, eval = curr_params$clustering_resolution != F}
cluster_markers <- xfun::cache_rds({
  
  cluster_markers <-
    Seurat::FindAllMarkers(seu, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
  cluster_markers
}, file = "cluster_markers.rds")
readr::write_tsv(cluster_markers, "cluster_markers.tsv")
```

# Celltype annotation

## SingleR

```{r singler_annots, eval = F, message = F, warning = F}
# get reference from celldex
singler_refs <- celldex::HumanPrimaryCellAtlasData() 

# get singler annots 
singler_annots <- xfun::cache_rds({
  singler_annots <- 
    SingleR::SingleR(
      test = SeuratWrappers::as.cell_data_set(seu, assay = "RNA")@assays@data$counts,
      ref = singler_refs,
      labels = singler_refs@colData$label.fine
    ) 
  singler_annots
}, file = "singler_annots.rds")

# main-to-reduced conversions
singler_main_to_reduced <-
  list(
    "Epithelial_cells" = c("Epithelial_cells", "Neuroepithelial_cell"),
    "Stromal_cells" = c("MSC", "Chondrocytes", "Osteoblasts"),
    "NK_cells" = c("NK_cell"),
    "B_cells" = c("B_cell", "Pre-B_cell_CD34-", "Pro-B_cell_CD34+"),
    "Erythrocytes" = c("Erythroblast", "Platelets", "MEP"),
    "Myeloid_cells" = c("Pro-Myelocyte", "Myelocyte", "Neutrophils", "Monocyte", "DC", "CMP", "GMP"),
    "Tissue_stem_cells" = c("Tissue_stem_cells", "iPS_cells", "Embryonic_stem_cells", "HSC_CD34+", "HSC_-G-CSF"),
    "Others" = c("Keratinocytes", "Hepatocytes", "Gametocytes", "BM & Prog.", "BM", "Neurons", "Astrocyte")
  ) %>%
  tibble::enframe(name = "label.reduced", value = "label.main") %>%
  tidyr::unnest(cols = "label.main")

# get reduced singler annots
singler_annots_reduced <- xfun::cache_rds({
  singler_annots_reduced <-
    tibble::tibble(label.fine = singler_annots$pruned.labels) %>%
    dplyr::left_join(
      singler_refs@colData %>%
        tibble::as_tibble() %>%
        dplyr::select(label.main, label.fine) %>%
        # fix many-to-many BM/BM & Prog. issue
        dplyr::filter(label.main != "BM & Prog.") %>%
        dplyr::distinct() 
    ) %>%
    dplyr::left_join(singler_main_to_reduced) %>%
    dplyr::mutate(
      label.reduced = dplyr::case_when(is.na(label.reduced) & is.na(label.main) ~ "Others",
                                       is.na(label.reduced) ~ label.main,
                                       TRUE ~ label.reduced))
  singler_annots_reduced
}, file = "singler_annots_reduced.rds")

# add to seurat object
seu <- xfun::cache_rds({
  seu@meta.data <-
    cbind(seu@meta.data,
          singler_annots_reduced)
  seu <- Seurat::SetIdent(seu, value = "label.reduced")
  seu
}, file = "seu_singler_annotated.rds")
```

```{r singler_reduced_plot}
p <-
  Seurat::DimPlot(seu, label = T , repel = T, label.size = 3, raster = F) &
  ditto_colours
p
```

## Malignancy and differentiation scores (from Zhang et al., 2021)

```{r malig_and_diff_scores}
# switch default assay to RNA
Seurat::DefaultAssay(seu) <- "RNA"
malignancy_genes <-
  readr::read_tsv(curr_params$annotate$malignancy_score_file) %>%
  dplyr::transmute(gene, 
                   population = dplyr::case_when(logFC > 0 ~ "malignant",
                                                 TRUE ~ "nonmalignant")) %>%
  {split(.$gene, .$population)} %>%
  {c(., list("differentiation" = c("KRT20", "PHGR1", "MDK", "CHDR2", "RARRES3", "GPA33", "SLC5A1", "MUC13")))}
seu <- 
  Seurat::AddModuleScore(
    seu, 
    assay = "RNA", 
    features = malignancy_genes, 
    name = names(malignancy_genes)
  )

# malignancy score = malignant score - nonmalignant score
seu$malignancy <- seu$malignant1 - seu$nonmalignant2

# differentiation score
seu$differentiation <- seu$differentiation3
seu$differentiation3 <- NULL
```

```{r malig_and_diff_plots, echo = F, fig.dim = c(10, 5)}
Seurat::FeaturePlot(seu, features = c("malignancy", "differentiation")) & 
  ggplot2::scale_colour_gradientn(colours = rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")))
```

```{r plot_malignacy_vs_nonmalignancy, echo = F, fig.dim = c(10, 5)}
seu@meta.data %>% 
  ggplot2::ggplot(ggplot2::aes(x = malignant1, y = nonmalignant2, colour = malignancy)) +
  ggplot2::geom_point() + 
  ggplot2::theme_classic() +
seu@meta.data %>%
  ggplot2::ggplot(ggplot2::aes(x = malignancy, y = differentiation)) +
  ggplot2::geom_point() + 
  ggplot2::theme_classic()
```

```{r save_annotated}
saveRDS(seu, "seu_annotated.rds")
```

## Canonical celltype markers

```{r avail_markers}
# load markers
markers <-
  curr_params$annotate$markers_file %>%
  readr::read_tsv(show_col_types = F) %>%
  {split(.$gene, .$population)}
# get markers present in the dataset
avail_markers <- markers %>% purrr::map(intersect, rownames(seu))
```

```{r canonical_marker_expr, eval = F}
# fig.dim = get_fig_dims(length(avail_markers), n_cols = 1)
p <- list()
names(avail_markers) %>%
purrr::map(function(ct) {
  seu %>%
    Nebulosa::plot_density(avail_markers[[ct]], joint = T) +
    ggplot2::ggtitle(paste(ct, paste(c(avail_markers[[ct]], ""), collapse = "+ "))) +
    patchwork::plot_layout(ncol = 1)
}) %>%
  setNames(names(avail_markers))
# create grob layout
p <-
  gridExtra::marrangeGrob(
    grobs = p,
    nrow = length(p),
    ncol = 1
  )
p
```


